---
title: "R data modes and structures"
output: pdf_document
---

# Data modes

Like any scripting language, `R`, has multiple data modes. Unlike many other languages, `R` does not require designation of a variable’s data mode prior to assignment. This is convenient, but also can cause problems when `R` assigns an unintended data mode to a variable. The common data modes in `R` include:

- numeric - real numbers (double) or integers

- complex - imaginary numbers

- logical - TRUE or FALSE; `R` also treats these as 1 and 0, respectively

- characer - strings or text values

# Data structures
Data of any of the modes described above is stored in one of `R`’s data structures. We will primarily use four data structures this semester: vectors, matrices/arrays, lists, and data frames.

## 1. Vectors: 
A vector is an `R` object that contains an ordered set of values. A vector is analogous to a row of data in Excel. Even variables assigned a single value are vectors of length 1. All values in a vector must be of the same data mode.

### a. Creating vectors

  - `c()`: combines specified values in a vector
  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=c(1,2,3,20)`
  
  - `:` : generates an ordered sequence incremented by 1
  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=1:4`

  - `seq()`: generates an ordered sequence incremented by the specified value or a sequence of the specified length
  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=seq(from=5,to=6,by=0.1)`
  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=seq(from=-10,to=-5,length.out=10)`

  - `rep()`: generates a vector of specified length containing the same value in each element of the vector

  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=rep(x=4,by=5)`

  - `numeric()`: generates a vector of specified length filled with 0’s
  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=numeric(length=5)`

  - `vector()`: generates a vector of specified length filled with FALSE’s
  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `v=vector(length=5)`

### b. Useful vector functions
  
  - mathematical operators: +, - , *, /

  - logical operators: <, >, <=, >=, ==, !=

  - `length()`: returns the length of the vector

  - `max()`: returns the maximum value contained in the vector

  - `min()`: returns the minimum value contained in the vector

  - `sum()`: returns the sum of the values in the vector

  - `cumsum()`: returns the cumulative sum for each element of the vector -mean(): returns the mean of the vector

  - `range()`: returns the minimum and maximum values

  - `var()`: returns the variance of the vector

  - `sd ()`: returns the standard deviation of the vector

  - `sort()`: returns a sorted version of the vector
  
  - `order()`: returns the numerical indices of vector elements in sorted order 
  
### c. Vector indexing and subsetting

Because vectors are an ordered list, a single element or subset of elements can be referred to using square brackets, [ ], and a numerical index.

```{r}
v=c(1,3,5,9,13)
v[1]
v[4]
v[c(1,3,5)]
v[-3]
```

An alternative means of indexing is a vector of logical values.

```{r}
v=c(1,3,5,9,13)
v>3
v[v>3]
```

The which() function creates numerical indices from a logical vector.

```{r}
v=c(1,3,5,9,13)
v>3
which(v>3)
```

The %in% function allows for logical indexing of overlapping sets.

```{r}
v=c(1,3,5,9,13)
w=seq(from=1,to=21,by=4)
v%in%w
v[!(v%in%w)]
```

Note that you cannot delete an element from a vector, but you can reassign a subset of a vector to the same variable.

```{r}
v=c(1,3,5,9,13)
v
v=v[v>3]
v
```

## 2. Matrices/Arrays: 
A matrix is a vector with two additional attributes, the number of rows and columns. Matrices are a special case (two-dimensional) of an array. Matrices are analogous to Excel worksheets and a three-dimensional array is analogous to an Excel workbook. Because `R` views arrays as vectors with some extra description about their shape, nearly all functions that work on a vector will work on a matrix/array. However, many functions will only work with matrices. All values in a matrix must be the same data mode.

### a. Creating matrices and arrays
  - `matrix()`: creates a matrix from a vector of values and dimensions

&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `A=matrix(0,nrow=2,ncol=2)`

  - `array()`: creates an array from a vector and dimensions

&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; `B=array(rep(1:4,2),dim=c(2,2,2))`

### b. Useful matrix functions

***many of these will work on higher dimensional arrays too
  
  - `dim()`: returns the dimensions (number of rows and columns) of the matrix

  - `nrow()`: returns the number of rows in the matrix

  - `ncol()`: returns the number of columns in the matrix

  - `rownames()`: returns the row names of the matrix; can also be used for assignment 
  
  - `colnames()`: returns the column names of the matrix; can also be used for assignment 
  
  - `rbind()`: add a vector to a specified matrix as a new row at the bottom of the matrix 
  
  - `cbind()`: add a vector to a specified matrix as a new column at the furthest right 
  
  - `%*%`: matrix multiplication

  - `t()`: transpose the matrix

  - `colMeans()`: calculate the mean of each column of the matrix

  - `colSums()`: calculate the sum of each column of the matrix

  - `apply()`: applies a function that works on a vector to each row or column of a matrix

### c. Matrix and array indexing/subsetting

Just like vectors, elements or subsets of matrices and arrays can be indexed using square brackets. Because an array is a vector a single number can be used to index a particular element, but this would require a lot of mental math to get the desired element. Instead, we can use a numerical index along each dimension of an array, separated by a comma, to index a particular element or subset of elements.

```{r}
M=matrix(1:4,nrow=2,ncol=2)
M[2,1]
M[2,2]
M[4]
M[,2]
```

Logical values work for indexing with matrices too.

```{r}
M=cbind(1:5,c(3,9,15,25,76))
M
M<8
M[M<8]
```

You can also subset a matrix based upon the content of a particular row or column.

```{r}
M=cbind(1:5,c(3,9,15,25,76))
M[M[,1]<3,]
M[M[,2]>24,1]
```

## 3. Lists: 

A list in `R` is somewhat like a vector, but can accommodate values of different data mode, including other data structures. One could create a list of matrices or even a list of lists. It also is often indexed by element names rather than a numerical index. `R`’s list data structure is very similar to a dictionary in `Python`.

### a. Creating lists

Lists are created using the `list()` function.

```{r}
a=list(name="walleye",length=225,tagged=TRUE)
a
```

### b. Useful list functions

  - `length()`: returns the number of components in the list

  - `names()`: returns the tags of a list

  - `lapply()`: analogous to `apply()`, but operates on lists and returns a list 
  
  - `sapply()`: the same functionality as `lapply()`, but returns a matrix or vector

### c. Indexing lists

There are a number of equivalent ways to access elements of a list, including element names or “tags” and square bracket (in this case double square brackets) indexing. For this reason, tags are optional, but tags do make referencing list components a lot easier.

```{r}
a=list(name="walleye",length=225,tagged=TRUE)
a$length
a[["length"]]
a[[2]]
```

### d. Adding and deleting list elements

Additional elements can easily be added to an existing list.

```{r}
a=list(name="walleye",length=225)
a
a$tagged=TRUE
a
```

Elements can be removed from a list be setting them equal to NULL.

```{r}
a=list(name="walleye",length=225,tagged=TRUE)
a$tagged=NULL
a
```

## IV. Data frames: 

A data frame is two-dimensional like a matrix, but can hold elements of different data modes. As matrices behave as vectors with a descriptor of shape, data frames behave as lists, but have two dimensions, which can often be very useful.

### a. Creating data frames

Data frames are created by the `data.frame()` function. If we want to maintain character data as characters we must use the argument "stringsAsFactors=FALSE" in the `data.frame()` function.

```{r}  
names=c("walleye","perch")
weight=c(270,57)
d=data.frame(names,weight,stringsAsFactors=FALSE)
d
```

### b. Useful data frame functions

  - `rbind()`: add a vector to a specified matrix as a new row at the bottom of the matrix 
  
  - `cbind()`: add a vector to a specified matrix as a new column at the furthest right
  
  - `colMeans()`: calculate the mean of each column of the matrix

  - `colSums()`: calculate the sum of each column of the matrix

  - `apply()`: applies a function that works on a vector to each row or column of a matrix 
  
  - `merge()`: joins two data frames together using a shared column as an index 
  
  - `lapply()`: analogous to `apply()`, but operates on lists and returns a list

  - `sapply()`: the same functionality as `lapply()`, but returns a matrix or vector
  
### c. Indexing data frames

Because data frames behave like matrices and lists we can index subsets and elements of data frames in almost any fashion.

*** Code continued from "a" above

```{r}
names=c("walleye","perch")
weight=c(270,57)
d=data.frame(names,weight,stringsAsFactors=FALSE)
d[[2]]
d[,2]
d$weight
```