---
title: "Lecture 06 - Git discussion, review, and extension"
output: 
  beamer_presentation:
    fig_caption: false
    includes:
      in_header: ../mypreamble.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## some logistics

- quizzes were worth too much and so I'm fixing that by increasing the weight on the group project

- messed up good & bads last week, but will do it on this Wednesday; put names on them!

- exercises due the following tutorial to allow for help at office hours

- orthogonal design to a point...

## Why use Git?

## How does Git work?

## How does Git work?

- A set of functions that allow tracking of progress in a project and facilitate collaboration.

- Sets up a hidden directory that stores information about changes in a working directory.

```{r, engine='bash', echo=TRUE, eval=FALSE}
  git init
  
  git add file_names
  git commit -m "custom commit message"
```

## Exercise 3 as a review

1. Use Git to commit each version of the shell script developed below when answering questions 2 - 4. Print your Git log for this shell script and turn it in with your answers below.

&nbsp;

2. Write a shell script that does NOT use a loop, but writes the atom number and element with the lowest value in the fifth column of each .pdb file to a text file. Once you have a shell script that works, write the contents of that script below.

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  cd data-shell/molecules
  
  git init
```

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  #!/usr/bin/env bash
  
  # a script for pulling the atom with the minimum value in column 5 from each of six .pdb files
  # usage ./pdbScript.sh
  
  cat cubane.pdb | grep ATOM | sort -n -k 5 | head -n 1 | 
  tr -s " " | cut -d " " -f 3,5 >> results.txt
  
  ...
  
  cat propane.pdb | grep ATOM | sort -n -k 5 | head -n 1 | 
  tr -s " " | cut -d " " -f 3,5 >> results.txt
```

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  git add -A
  
  git commit -m "added shell script without a for loop"
```

## Exercise 3 as a review

3. Revise the shell script from above so that it DOES use a loop. Once you have a shell script that works, write the contents of that script below.

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  #!/usr/bin/env bash
  
  # a script for pulling the atom with the minimum value in column 5 from each of six .pdb files
  # usage ./pdbScript.sh
  
  for file in "$@"
  do
    cat $file | grep ATOM | sort -n -k 5 | head -n 1 | 
    tr -s " " | cut -d " " -f 3,5 >> results.txt
  done
```

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  git add -A
  
  git commit -m "added a for loop to pdbScript.sh"
```
## Exercise 3 as a review

4. Revise the shell script from above so that it can take an argument from the user to determine which column from the .pdb files we are using to pick an atom. Once you have a shell script that works, write the contents of that script below.

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  git log
  
  git diff HEAD 'hash from previous commit'
  
  git checkout 'hash from previous commit'
  -- pdbScript.sh
```

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  #!/usr/bin/env bash
  
  # a script for pulling the atom with the minimum value in column 5 from each of six .pdb files
  # usage ./pdbScript.sh column_number
  
   cat cubane.pdb | grep ATOM | sort -n -k $1 | head -n 1 | 
  tr -s " " | cut -d " " -f 3,5 >> results.txt
  
  ...
  
  cat propane.pdb | grep ATOM | sort -n -k $1 | head -n 1 | 
  tr -s " " | cut -d " " -f 3,5 >> results.txt
```

## Exercise 3 as a review

```{r, engine='bash', echo=TRUE, eval=FALSE}
  git add -A
  
  git commit -m "added argument to pbsScript.sh 
  for selecting the column number but got rid of 
  the for loop"
```


## What is the connection between Git and GitHub?

## What is the connection between Git and GitHub?

\begin{figure}
  \includegraphics[width=0.85\textwidth]{GitHub.jpg}
\end{figure}

## Sign up for GitHub...

\begin{figure}
  \includegraphics[width=\textwidth]{GitHubSignup.jpg}
\end{figure}

## Sign up for GitHub...

\begin{figure}
  \includegraphics[width=\textwidth]{GitHubLandingpage.jpg}
\end{figure}

## Sign up for GitHub...

\begin{figure}
  \includegraphics[width=\textwidth]{GitHubProfile.jpg}
\end{figure}

## function review

`git config`

`git init`

`git add`

`git commit`

`git status`

`git log`

`git checkout`