---
title: "Lecture 04 - Unix loops"
output: 
  beamer_presentation:
    fig_caption: false
    includes:
      in_header: ../mypreamble.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## for loops in general

**for** variable **in** set

&nbsp; &nbsp; code to be executed

&nbsp;

- variable: a place holder used to temporarily hold each element in the set

- set: the entities you want to have the code within the for loop executed for

## for loops in Unix
`for variable in set`

`do`

&nbsp; &nbsp; *code to be executed using* `$variable`

`done`

## potato example
```{r, engine='bash', echo=TRUE, eval=TRUE}
for number in one two
do
  echo $number potato
done
```

## Executing a for loop at the Unix command line

- usually the Unix command prompt is a `$`

- once you've begun a for loop the prompt changes to `>`

- this is only to remind you that you are in a loop

- you can get out of the loop or any running command with *Ctrl-c*

## Reminder again

Most often the variables we will include in a set are file names, therefore you should remember...

&nbsp;

- NO SPACES IN FILE NAMES!!!!!

&nbsp;

- File extensions don't mean anything to Unix!

## By the way...

Did anyone else get files with name stats-stats-filename when working through tutorial?

&nbsp;

Why?

&nbsp;

How to avoid this?

## More than one shell window!!!

Holy cows, you can do this?

&nbsp;

Remember multi-processing!

## Echo and debugging

Why is `echo` useful for testing `for` loops?

## Echo and debugging

Why is `echo` useful for testing `for` loops?

&nbsp;

```{r, engine='bash', echo=TRUE, eval=TRUE}
  for file in *.txt
  do
    echo "bash goostats $file stats-$file"
  done
```

## What happened on a process level?

```{r, engine='bash', echo=TRUE, eval=FALSE}
  for file in *.txt
  do
    echo "bash goostats $file stats-$file"
  done
```

\begin{figure}
    \includegraphics[width=0.98\textwidth]{process.jpg}
\end{figure}

## for loop application

**Challenge:** Use the files in `molecules/` to print each # of atoms:

`cubane:`

&nbsp; &nbsp; &nbsp; &nbsp; `16`

`ethane:`

&nbsp; &nbsp; &nbsp; &nbsp; `8`

`methane:`

&nbsp; &nbsp; &nbsp; &nbsp; `5`

`octane:`

&nbsp; &nbsp; &nbsp; &nbsp; `26`

`pentane:`

&nbsp; &nbsp; &nbsp; &nbsp; `17`

`propane:`

&nbsp; &nbsp; &nbsp; &nbsp; `11`

## for loop application

```{r, engine='bash', echo=TRUE, eval=FALSE}
  for file in *.pdb
  do
    echo $file | sed 's/.pdb/: /'
    cat $file | grep ATOM | wc -l
  done
```

## storing for loop output in a file

```{r, engine='bash', echo=TRUE, eval=FALSE}
  for file in *.pdb
  do
    echo $file | sed 's/.pdb/: /' > output.txt
    cat $file | grep ATOM | wc -l > output.txt
  done
```

**VS**

```{r, engine='bash', echo=TRUE, eval=FALSE}
  for file in *.pdb
  do
    echo $file | sed 's/.pdb/: /' >> output.txt
    cat $file | grep ATOM | wc -l >> output.txt
  done
```

## nested for loops

- As shown in the tutorial, this is useful if you have multiple directories with similar files

&nbsp;

```{r, engine='bash', echo=TRUE, eval=FALSE}
  for dir in */
  do
    for file in *.txt
    do
      echo "$file is in $dir"
    done
  done
```

## function review
`for`

`$`

`history`

*up arrow*

*Ctrl-c*

*Ctrl-a*

*Ctrl-e*