---
title: "Lecture 13 - Implementing A Layered Grammar of Graphics using R and ggplot2"
output: pdf_document
---
```{r,echo=FALSE}
knitr::opts_chunk$set(fig.width=3,fig.height=2,tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

\pagenumbering{gobble}

## Tutorial

Most of today will be spent working through examples of a variety of plot types that can be generated using R and ggplot2. Work through the provided code and talk to your neighbor about what is happening and why it works. Make sure to ask any questions you may have.

&nbsp;

```{r}
library(ggplot2)
library(grid)
library(gridExtra)

mpg=read.table("mpg.txt",header=TRUE,sep="\t",stringsAsFactors=FALSE)

dim(mpg)

head(mpg)

# plot of displacement (engine size) vs. city miles per gallon (cty)
a=ggplot(data=mpg,aes(x=displ,y=cty))
a+geom_point()+coord_cartesian()

# remove grey background
a+geom_point()+coord_cartesian()+theme_bw()

# remove grey background and gridlines
a+geom_point()+coord_cartesian()+theme_classic()

# change the x and y labels; cartesian coordinates are default
a+geom_point()+theme_classic()+xlab("displacment (l)")+ylab("miles per gallon-city")

# arguments to geom_point() can alter the appearance of the points
a+geom_point(color="blue",shape=22,size=3)+theme_classic()+xlab("displacment (l)")+ylab("miles per gallon-city")

# log transform the y-axis
a+geom_point()+theme_classic()+xlab("displacment (l)")+ylab("miles per gallon-city")+scale_y_log10()

# arguments to scale can also customize the range and tick locations
a+geom_point()+theme_classic()+xlab("displacment (l)")+ylab("miles per gallon-city")+scale_y_log10(limits=c(1,100),breaks=c(1,10,100))

# we can also color code points based on continuous or categorical variables
# continuous
a+geom_point(aes(color=cyl))+theme_classic()+xlab("discplacement (l)")+ylab("miles per gallon-city")
# categorical
a+geom_point(aes(color=as.factor(cyl)))+theme_classic()+xlab("discplacement (l)")+ylab("miles per gallon-city")
# categorical - the default colors are a bit odd
a+geom_point(aes(color=as.factor(cyl)))+theme_classic()+xlab("discplacement (l)")+ylab("miles per gallon-city")+scale_color_manual(values=c('red','green','blue','orange'))+theme(legend.title=element_blank())

# add a linear trendline with a new layer
a+geom_point()+theme_classic()+xlab("displacment (l)")+ylab("miles per gallon-city")+stat_smooth(method="lm")

# add a spline with a new layer
a+geom_point()+theme_classic()+xlab("displacment (l)")+ylab("miles per gallon-city")+stat_smooth(method="loess")


# histogram of mpg hwy
b=ggplot(data=mpg,aes(x=hwy))
b+geom_histogram()+theme_classic()

# change bins
b+geom_histogram()+theme_classic()+stat_bin(binwidth=5)

# change bins; the stat_bin is a default for geom_histogram, so it can be skippped
# color can also be specified in the geom_histogram call
b+geom_histogram(binwidth=5,fill='red',color='black')+theme_classic()


# we can generate a barplot of means too
d=ggplot(data=mpg)
d+geom_bar(aes(x=as.factor(drv),y=cty),stat="summary",fun.y="mean",fill='black',color='black')+theme_classic()+xlab("drive")+ylab("miles per gallon city")

tapply(X=mpg$cty,INDEX=mpg$drv,FUN=mean)


# faceting allows the same plot for different classes to be generated
f=ggplot(data=mpg,aes(x=displ,y=cty))+geom_point()
f+facet_wrap(~drv)+theme_classic()

# grid.arrange from grid/gridExtra libraries allows for multiple panels with different plot types
# you'll need to install the grid and gridExtra libraries to accomplish this!
g=ggplot(data=mpg,aes(x=displ,y=cty))+geom_point(aes(color=as.factor(cyl)))+theme_classic()+xlab("displacement (l)")+ylab("miles per gallon-city")+scale_color_manual(values=c('red','green','blue','orange'))+theme(legend.title=element_blank())

h=ggplot(data=mpg)+geom_bar(aes(x=as.factor(drv),y=cty),stat="summary",fun.y="mean")+theme_classic()

grid.arrange(g,h,ncol=2)
```

## Challenge

Practice using the syntax demonstrated above by writing a script to generate the following plots using the mpg data.

1\. A scatter plot of miles per gallon city versus miles per gallon highway. Color code the points by 'drv' (four-wheel drive vs. front-wheel drive vs. rear-wheel drive). Add a linear trendline to the plot.

&nbsp;

2\. A "density plot" of engine displacement.

&nbsp;

3\. A barplot of mean displacement for different numbers of cylinders (cyl).