---
title: "Lecture 05 - Unix shell scripts"
output: 
  beamer_presentation:
    fig_caption: false
    includes:
      in_header: ../mypreamble.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## why write a shell script?

## why write a shell script?

- repeat tasks easily

- reproducible and documented (comments!!!!)

- modularize your analysis processes

## arguments to a shell script

- `$1`, `$2`, `$3`, etc.

&nbsp;

- `$@` - useful with loops in a shell script!

## shell scripts in an 'orthoganal' manner

- remember the power of Unix is it's modularity

&nbsp;

- BUILD SCRIPTS TO READ FROM `stdin` AND WRITE TO `stdout`

## convert a SWC shell script to use `stdin` and `stdout`

```{r, engine='bash', echo=TRUE, eval=FALSE}
  # Select lines from the middle of a file.
  # Usage: bash middle.sh filename end_line num_lines
  head -n "$2" "$1" | tail -n "$3"
```

`bash middle.sh pentane.pdb 15 5`

## convert a SWC shell script to use `stdin` and `stdout`

```{r, engine='bash', echo=TRUE, eval=FALSE}
  # Select lines from the middle of a file.
  # Usage: middle2.sh end_line num_lines
  head -n "$1" | tail -n "$2"
```

`cat pentane.pdb | bash middle2.sh 15 5`

## making scripts behave even more like a function

To behave like other basic functions a shell script needs:

- a 'shebang' as the first line

    + `#!/path/to/interpretor`

    + this tells the shell what to use to run the text in the script

&nbsp;

- the script file must be executable by the user

## file permissions

if we use `ls -l *.sh` to look at our shell scripts, we can see read, write, and execute permissions

- usually the first character indicates files vs. directories

&nbsp;

- the next nine characters indicate the presence or absence of read (r), write (w), and execute (x) permissions for the "owner", "group", and "other" users

## changing file permissions

- the function `chmod` can be used to change permissions

&nbsp;

- to add executable permission for the owner try: 
`chmod u+x filename`

    + *who* comes first: user (u), group (g), other (o), or all (a)

    + "op" comes next: this indicates if we are adding (+) or removing (-) permissions
    
    + "permission" is last: read (r), write (w), and execute (x) are the most common

## convert a SWC shell script to use `stdin` and `stdout`

```{r, engine='bash', echo=TRUE, eval=FALSE}
  #!/usr/bin/env bash
  
  # Select lines from the middle of a file.
  # Usage: middle3.sh end_line num_lines
  head -n "$1" | tail -n "$2"
```

`chmod u+x middle3.sh`

`cat pentane.pdb | ./middle3.sh 15 5`

## convert a SWC shell script to use `stdin` and `stdout`

`bash middle.sh pentane.pdb 15 5`

&nbsp;

`cat pentane.pdb | bash middle2.sh 15 5`

&nbsp;

`cat pentane.pdb | ./middle3.sh 15 5`

## version control and Git

A computational best practice that solves this problem, among others...

\begin{figure}
  \includegraphics[width=0.5\textwidth]{phd101212s.jpg}
\end{figure}


## version control and Git

- A set of functions that allow tracking of progress in a project and facilitate collaboration.

- Becoming the gold standard in biocomputation-focused research.

- Sets up a hidden directory that stores information about changes in a working directory.

```{r, engine='bash', echo=TRUE, eval=FALSE}
  git init
  
  git add file_names
  git commit -m "custom commit message"
```

- Consult documentation ('Git_Setup.pdf') available on Sakai for installation instructions.

## function review

`chmod`

&nbsp;

`#!/path/to/interpretor`