---
output: pdf_document
---

\pagenumbering{gobble}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quiz 04 - Unix shell scripts

1\) Why would one use a shell script?

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

2\) If one had the following, poorly commented, shell script (`middle.sh`), fill in the blanks to execute the script correctly on the file `input.txt` and put 10 lines into `output.txt`:

```{r, engine='bash', echo=TRUE, eval=FALSE}
  # Select lines from the middle of a file.
  cat $1 | head -n 25 | tail -n $3 > $2 
```

`$bash middle.sh` _______________   ________________   ________________

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

## Quiz 04 - Unix shell scripts

1\) Why would one use a shell script?

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

2\) If one had the following, poorly commented, shell script (`middle.sh`), fill in the blanks to execute the script correctly on the file `input.txt` and put 10 lines into `output.txt`:

```{r, engine='bash', echo=TRUE, eval=FALSE}
  # Select lines from the middle of a file.
  cat $1 | head -n 25 | tail -n $3 > $2 
```

`$bash middle.sh` _______________   ________________   ________________

&nbsp;

&nbsp;

